---
title: "R4ds Chapter 5 Data Transformation Yuqing Xue Personal Note"
author: "Yuqing Xue"
date: "October 22, 2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

<br>

Notes and code follow Hadley Wickham's 'R for Data Science' material, Chapter 5. <http://r4ds.had.co.nz/transform.html#introduction-2>

### 5.1 Introduction


#### `nycflights13`
```{r message=FALSE, warning=FALSE}
##########################################
#######  5 Data Transformation
########################################## 
library(nycflights13)
library(tidyverse)
nycflights13::flights
```

<br>

### 5.2 Filter

#### Explore, try `%in%`
```{r explore}

jan1 <- filter(flights, month == 1, day == 1)

nov_and_dec = filter(flights, month == 11 | month == 12)
## A useful short-hand for this problem is x %in% y. This will select every row where x is one of the values in y.
nov_dec <- filter(flights, month %in% c(11, 12))
```

```{r NYC to San Fransisco Bay Area}

NYCtoSF <-  
  flights %>% 
    filter(dest %in% c("SFO", "OAK", "SJC")) %>%
      select(year, month, day, carrier, origin, dest, air_time)
NYCtoSF

NYCtoSF2 <-  
  flights %>% 
    filter(dest == "SFO" | dest == "OAK" |  dest == "SJC") %>%
      select(year, month, day, carrier, origin, dest, air_time)
NYCtoSF2

# H.W. As well as & and |, R also has && and ||. Don’t use them here! You’ll learn when you should use them in conditional execution.

``` 

<br>

####  5.2.4 Exercises
```{r exercise 1.1}
 
#################################
### 1. Find all flights that
# ###############################
## 1. Had an arrival delay of two or more hours
## 2. Flew to Houston (IAH or HOU)
## 3. Were operated by United, American, or Delta
## 4. Departed in summer (July, August, and September)
## 5. Arrived more than two hours late, but didn’t leave late
## 6. Were delayed by at least an hour, but made up over 30 minutes in flight
## 7. Departed between midnight and 6am (inclusive)

Ex1.1 <- 
  flights %>%
    filter(arr_delay >= 120) %>%
      arrange(desc(arr_delay)) %>%
        select(month, day, carrier, origin, dest, dep_delay, arr_delay)
Ex1.1
```

```{r exercise 1.2 - 1.4}
Ex1.2 = filter(flights, dest %in% c("IAH","HOU"))
Ex1.3 = filter(flights, carrier %in% c("UA","AA","DL"))
Ex1.4 = filter(flights, month %in% c(7,8,9))
```

```{r exercise 1.5 - 1.7}
Ex1.5 = filter(flights, arr_delay >= 120 & dep_delay <= 0 )
Ex1.6 = filter(flights, dep_delay >= 60 & dep_delay - arr_delay >= 30)
Ex1.7 = filter(flights, dep_time <= 600)
```

```{r exercise 2}

#################################
### 2. Another useful dplyr filtering helper is between(). 
###    What does it do? Can you use it to simplify the code needed to answer the previous challenges?
#################################

?between
Ex1.4b = filter(flights, between(month, 7, 9))
```


```{r exercise 3}
#################################
### 3. How many flights have a missing dep_time? What other variables are missing? 
###    What might these rows represent?
#################################

sum(is.na(flights$dep_time)) # 8255

flights_missing_dep_time <- flights %>% filter(is.na(dep_time))
flights_missing_dep_time 

flights_cancelled <- flights %>% 
  filter(is.na(dep_time) & is.na(dep_delay) & is.na(arr_time) & is.na(arr_delay) & is.na(arr_time))
# same 8255 flights
```

Along with dep_time, dep_delay, arr_time, arr_delay and air_time are also missing, for all 8255 flights. It probably means that those flights were cancelled.

```{r exercise 4}

##################################################
### 4. Why is NA ^ 0 not missing? Why is NA | TRUE not missing? 
### Why is FALSE & NA not missing? Can you figure out the general rule? (NA * 0 is a tricky counterexample!)
##################################################

```


### 5.3 Arrange

`arrange()` works similarly to `filter()` except that instead of selecting rows, it changes their order. It takes a data frame and a set of column names (or more complicated expressions) to order by. If you provide more than one column name, each additional column will be used to break ties in the values of preceding columns:

```{r}
arrange(flights, year, month, day)
arrange(flights, desc(arr_delay))
```

```{r}
df <- tibble(x = c(5, 2, NA))
arrange(df, x)
```
In this way, `x` then becomes the column name of the tibble `df`.

#### 5.3.1 Exercise
##### 1. How could you use `arrange()` to sort all missing values to the start? (Hint: use `is.na()`).

```{r}
arrange(df, !is.na(x))
```
By the logic that ' `FALSE` is less than `TRUE` '.



